Week 7 Spatial Autocorrelation

Analysing spatial autocorrelation with Moran's I, LISA etc.

```{r}
library(here)
library(janitor)
library(sf)
library(tidyverse)
library(usethis)
library(tmap)

# read the ward data in
Londonwards <- st_read(here::here("week7_practical_data", "London-wards-2018_ESRI", "London_ward.shp"))

# inspect
qtm(Londonwards)

```

```{r}
LondonWardsMerged <- st_read(here::here("week7_practical_data", "statistical-gis-boundaries-london", "ESRI", "London_Ward_CityMerged.shp")) %>% 
  st_transform(., 27700)

#inspect
qtm(LondonWardsMerged)
```

```{r}
WardData <- read_csv("https://data.london.gov.uk/download/ward-profiles-and-atlas/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv", locale = locale(encoding = "latin1"), na = c("NA", "n/a")) %>% 
  clean_names()

LondonWardsMerged <- LondonWardsMerged %>% left_join(WardData, by = c("GSS_CODE" = "new_code")) %>% 
  dplyr::distinct(GSS_CODE, .keep_all = T) %>% 
  dplyr::select(GSS_CODE, ward_name, average_gcse_capped_point_scores_2014)

View(LondonWardsMerged)
```

```{r}
# have a look at the crs to check if correct
st_crs(LondonWardsMerged)
```

```{r}
BluePlaques <- st_read(here::here("week7_practical_data", "open-plaques-london-2018-04-08.geojson")) %>% 
  st_transform(., 27700)

```

```{r}
tmap_mode("plot")
tm_shape(LondonWardsMerged) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(BluePlaques) +
  tm_dots(col = "blue")
```

Data Cleaning

```{r}
summary(BluePlaques)
```

Need to lose the blue plaques that fall outisde of London

```{r}
BluePlaquesSub <- BluePlaques[LondonWardsMerged,]

tm_shape(LondonWardsMerged) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(BluePlaquesSub) +
  tm_dots(col = "blue")
```

Data Manipulation

Take the length of each list per polygon and add this as a new column

```{r}
points_sf_joined <- LondonWardsMerged %>% 
  mutate(n = lengths(st_intersects(., BluePlaquesSub))) %>% 
           janitor::clean_names() %>% 
           # calculate area
           mutate(area = st_area(.)) %>% 
           # then density of the points per ward
           mutate(density = n / area) %>% 
           # select density and some other variables
           dplyr::select(gss_code, ward_name, n, density, average_gcse_capped_point_scores_2014)

# inspect
View(points_sf_joined)
```

```{r}
# quick choropleth map to check how we are getting on
points_sf_joined <- points_sf_joined %>% 
  group_by(gss_code) %>% 
  summarise(density = first(density),
            wardname = first(ward_name),
            plaquecount = first(n))

View(points_sf_joined)

tm_shape(points_sf_joined) +
  tm_polygons("density",
              style = "jenks",
              palette = "PuOr",
              midpoint = NA,
              popup.vars = c("wardname", "density"),
              title = "Blue Plaque Density")
```

```{r}
library(spdep)
```

```{r}
# first calculate the centroids of all wards in London
coordsW <- points_sf_joined %>% 
  st_centroid() %>% 
  st_geometry()

plot(coordsW, axes = TRUE)
```

Now we need to generate a spatial weights matrix

```{r}
# create a neighbours list
LWard_nb <- points_sf_joined %>% 
  poly2nb(., queen = T)

# inspect
summary(LWard_nb)
```

```{r}
# plot
plot(LWard_nb, st_geometry(coordsW), col = "red")
# add a map underneath
plot(points_sf_joined$geometry, add = T)
```

```{r}
# create a spatial weights matrix from these weights. B means binary here
Lward.lw <- LWard_nb %>% 
  nb2mat(., style="B")

# we have 3,680 neighbours
sum(Lward.lw)
```

```{r}
# Moran's I requires a spatial weight list type object as opposed to matrix
Lward.lw <- LWard_nb %>% 
  nb2listw(., style = "C")
```

Moran's I test tells us whether we have clustered values (close to 1) or dispersed values (close to -1).

We will calculate for the densities rather than raw values (Why? - raw number depends greatly on ward size and population etc. Density matters more to compare wards amongst one another)

```{r}
I_LWard_Global_Density <- points_sf_joined %>% 
  pull(density) %>% 
  as.vector() %>% 
  moran.test(., Lward.lw)

I_LWard_Global_Density
```

Geary's C tells us whether similar values or dissimilar values are clustering

```{r}
C_LWard_Global_Density <- 
  points_sf_joined %>% 
  pull(density) %>% 
  as.vector() %>% 
  geary.test(., Lward.lw)

C_LWard_Global_Density
```

Getis Ord General G tells us whether high or low values are clustering. G > Expected = high values clustering vs G < Expected = low values clustering

```{r}
G_LWard_Global_Density <- 
  points_sf_joined %>% 
  pull(density) %>% 
  as.vector() %>% 
  globalG.test(., Lward.lw)

G_LWard_Global_Density
```

Moran's I statistic = 0.67 meaning we have some distinctive clustering

Geary's C statistic = 0.41 meaning that similar values are clustering

General G statistic = G > Expected, so high values are tending to cluster

Local Moran's I

```{r}
# use the localmoran function to generate I for each ward in the city

colnames(points_sf_joined)

I_LWard_Local_count <- points_joined_summary %>% 
  pull(plaquecount) %>% 
  as.vector() %>% 
  localmoran(., Lward.lw) %>% 
  as_tibble()

I_LWard_Local_Density <- points_joined_summary %>% 
  pull(density) %>% 
  as.vector() %>% 
  localmoran(., Lward.lw) %>% 
  as_tibble()

# what does that output look like?
slice_head(I_LWard_Local_Density, n=10)
```

We want to copy some of the columns (the I score and the z-score standard deviation column back into the LondonWards spatialPolygonsDataframe)

```{r}
points_sf_joined <- points_sf_joined %>% 
  mutate(plaque_count_I = as.numeric(I_LWard_Local_count$Ii)) %>% 
  mutate(plaque_count_Iz = as.numeric(I_LWard_Local_count$Z.Ii)) %>% 
  mutate(density_I = as.numeric(I_LWard_Local_Density$Ii)) %>% 
  mutate(density_Iz = as.numeric(I_LWard_Local_Density$Z.Ii))
```

Mapping

We'll set the breaks manually based on the rule that data points >2.58 or <-2.58 standard deviations away from the mean are significant at the 99% level (<1% chance that autocorrelation is not present)

```{r}
breaks1 <- c(-1000, -2.58, -1.96, -1.65, 1.65, 1.96, 2.58, 1000)
```

rev() function makes the higher values correspond to red

```{r}
library(RColorBrewer)
MoranColours <- rev(brewer.pal(8, "RdGy"))

View(points_sf_joined)
```

Plot an interactive map

```{r}
tm_shape(points_sf_joined) +
  tm_polygons("plaque_count_Iz",
              style = "fixed",
              breaks = breaks1,
              palette = MoranColours,
              midpoint = NA,
              title = "Local Moran's I, Blue Plaques in London")
```

Local Getis Ord G

The result of Getis Ord G analysis is an array of Z-scores, one for each pixel [or polygon], which is the number of standard deviations that the pixel [or polygon] and its neighbors are from the global mean. High Z-scores indicate more intense clustering of high pixel values, indicating hot spots. Low Z-scores indicate more intense clustering of low values, indicating cold spots. Individual pixels with high or low values by themselves might be interesting but not necessarily significant.

```{r}
Gi_LWard_Local_Density <- points_sf_joined %>% 
  pull(density) %>% 
  as.vector() %>% 
  localG(., Lward.lw)

head(Gi_LWard_Local_Density)
```

Add the Getis Ord G data to the simple feature

```{r}
points_sf_joined <- points_sf_joined %>% 
  mutate(density_G= as.numeric(Gi_LWard_Local_Density))
```

And map the outputs

```{r}
GIColours <- rev(brewer.pal(8, "RdBu"))

# now plot on an interactive map
tm_shape(points_sf_joined) +
  tm_polygons("density_G",
              style = "fixed",
              breaks = breaks1,
              palette = GIColours,
              midpoint = NA,
              title = "Gi*, Blue-Plaques in London")
```

Other variables - let's have a look at average GCSE scores

```{r}
# use htead to see what other variables are in the data file
slice_head(points_sf_joined, n = 2)
View(points_sf_joined)
```

```{r}
# Or print out the class of each column
Datatypelist <- LondonWardsMerged %>% 
  st_drop_geometry() %>% 
  summarise_all(class) %>% 
  pivot_longer(everything(),
               names_to = "All_variables",
               values_to = "Variable_class")

Datatypelist
```

```{r}
I_LWard_Local_GCSE <- LondonWardsMerged %>% 
  arrange(GSS_CODE) %>% 
  pull(average_gcse_capped_point_scores_2014) %>% 
  as.vector() %>% 
  localmoran(., Lward.lw) %>% 
  as_tibble()

points_sf_joined <- points_sf_joined %>% 
  arrange(gss_code) %>% 
  mutate(GCSE_LocIz = as.numeric(I_LWard_Local_GCSE$Z.Ii))

points_sf_joined

tm_shape(points_sf_joined) +
  tm_polygons("GCSE_LocIz",
              style = "fixed",
              breaks = breaks1,
              palette = MoranColours,
              midpoint = NA,
              title = "Local Moran's I, GCSE Scores")
```

Now the Gi* statistic to look at clusters of high and low scores and explain what the output map is showing

```{r}
G_LWard_Local_GCSE <- LondonWardsMerged %>% 
  dplyr::arrange(GSS_CODE) %>% 
  dplyr::pull(average_gcse_capped_point_scores_2014) %>% 
  as.vector() %>% 
  localG(., Lward.lw)

points_sf_joined <- points_sf_joined %>% 
  dplyr::arrange(gss_code) %>% 
  dplyr::mutate(GCSE_LocGiz = as.numeric(G_LWard_Local_GCSE))

tm_shape(points_sf_joined) +
  tm_polygons("GCSE_LocGiz",
              style = "fixed",
              breaks = breaks1,
              palette = GIColours,
              midpoint = NA,
              title = "Gi* GCSE Scores")
```

End of the practical